Population scale mapping of transposable element diversity reveals links to gene regulation and epigenomic variation

### 2016-08-18 Paper/paper.tex
Updates to text

\usepackage[scaled]{helvet}
\renewcommand\familydefault{\sfdefault}
\usepackage[T1]{fontenc}

@@

Variation in the presence or absence of transposable elements (TEs) is
a major source of genetic variation between individuals. Here, we
identified 23,095 TE presence/absence variants between 216 Arabidopsis
accessions. Most TE variants were rare, and we find a burden of rare
variants associated with local extremes of both gene expression and
DNA methylation levels within the population. Of the common alleles
identified, two thirds were not in linkage disequilibrium with nearby
SNPs, implicating these TE variants as a source of novel genetic
diversity. Nearly 200 common TE variants were associated with
significantly altered expression of nearby genes, and a major fraction
of inter-accession DNA methylation differences were associated with
nearby TE insertion variants. Overall, we find that TE variants likely
play an important role in facilitating epigenomic and transcriptional
variation between wild Arabidopsis accessions, indicating a strong
genetic basis for epigenetic variation.

@@

Transposable elements (TEs) are mobile genetic elements present in
nearly all studied organisms, and comprise a large fraction of most
eukaryotic genomes. The two classes of TEs are retrotransposons (class
I elements), which transpose via an RNA intermediate requiring a
reverse transcription reaction, and DNA transposons (class II
elements), which transpose via either a cut-paste or, in the case of
Helitrons, a rolling circle mechanism with no RNA intermediate
\cite{Wicker:2007en}. TE activity poses mutagenic potential as a TE
insertion may disrupt functional regions of the genome. Consequently,
safeguard mechanisms have evolved to suppress this activity, including
the methylation of cytosine nucleotides (DNA methylation) to produce
5-methylcytosine (mC), a modification that can induce transcriptional
silencing of the methylated locus. In \emph{Arabidopsis thaliana}
(Arabidopsis), DNA methylation occurs in all three DNA sequence
contexts: mCG, mCHG, and mCHH, where H is any base but
G. Establishment of DNA methylation marks can be carried out by two
distinct pathways -- the RNA-directed DNA methylation pathway guided
by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2 pathway
\cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
methylation in Arabidopsis is in the transcriptional silencing of TEs.
Loss of DNA methylation due to mutations in genes essential for its
establishment or maintenance leads to expression of previously silent
TEs, and in some cases transposition \cite{Mirouze:2009km,Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu,Zemach:2013dj}.

@@

The tools developed previously for identification of novel TE insertion
events have several limitations. They either require a library of active
TE sequences, cannot identify TE absence variants, are not designed
with population studies in mind, or suffer from a high degree of
false-negatives
\cite{Thung:2014ir, Robb:2013gw, Henaff:2015dl, Quadrana:2016bi}. In
order to accurately map the locations of TE presence/absence variants
with respect to a reference genome, we have developed a novel algorithm,
TEPID (Transposable Element Polymorphism IDentification), which is
designed for population studies. We tested our algorithm using both
simulated and real Arabidopsis sequencing data, finding that TEPID is
able to accurately identify TE presence/absence variants with respect to
the Col-0 reference genome. We applied our TE analysis method to
existing genome resequencing data for 216 different wild Arabidopsis
accessions, and identified widespread TE variation amongst these
accessions \cite{Schmitz:2013iu}.

@@

We developed TEPID, an analysis pipeline capable of detecting TE
presence/absence variants from paired end DNA sequencing data. TEPID
integrates split and discordant read mapping information, read mapping
quality, sequencing breakpoints, as well as local variations in
sequencing coverage to identify novel TE presence/absence variants with
respect to a reference TE annotation (Figure \ref{fig1}; see methods). This
typically takes 5-10 minutes using Arabidopsis sequencing data at 20-40x
coverage (excluding the read mapping step). After TE variant discovery
has been performed, TEPID then includes a second refinement step
designed for population studies. This examines each region of the genome
where there was a TE insertion identified in any member of the group,
and checks for evidence of this insertion in all other members of the
group. In this way, TEPID leverages TE variant information for a group
of related samples to correct false negative calls within the group.
Testing of TEPID using simulated TE variants in the Arabidopsis genome
showed that it was able to reliably detect simulated TE variants at
sequencing coverage levels commonly used in genomics studies (Figure 1 -
figure supplement \ref{fig1s1}).

In order to further assess the sensitivity and specificity of TE variant
discovery using TEPID, we identified TE variants in the Landsberg
\emph{erecta} (L\emph{er}) accession, and compared these with the
L\emph{er }genome assembly created using long PacBio sequencing reads
\cite{Chin:2013iw}. Previously published 100 bp paired-end L\emph{er}
genome resequencing reads \cite{Schneeberger:2011ft} were first analyzed
using TEPID, enabling identification of 446 TE insertions (Figure 1 -
source data 1) and 758 TE absence variants (Figure 1 - source data 2)
with respect to the Col-0 reference TE annotation. Reads providing
evidence for these variants were then mapped to the L\emph{er} reference
genome, generated by \emph{de novo} assembly using Pacific
Biosciences P5-C3 chemistry with a 20 kb insert library
\cite{Chin:2013iw}, using the same alignment parameters as was used to
map reads to the Col-0 reference genome. This resulted in 98.7\% of
reads being aligned concordantly to the L\emph{er} reference, whereas
100\% aligned discordantly or as split reads to the Col-0 reference
genome (Table \ref{table1}). To find whether reads mapped to homologous regions in
both the Col-0 and L\emph{er} reference genomes, we conducted a blast
search \cite{Camacho:2009fc} using the DNA sequence between read pair
mapping locations in the L\emph{er }genome against the Col-0 genome, and
found the top blast result for 80\% of reads providing evidence for TE
insertions, and 89\% of reads providing evidence for TE absence variants
in L\emph{er}, to be located within 200 bp of the TE variant
reported by TEPID. We conclude that reads providing evidence for TE
variants map discordantly or as split reads when mapped to the Col-0
reference genome, but map concordantly to homologous regions of the
L\emph{er} \emph{de novo} assembled reference genome, indicating that
structural variation is present at the sites identified by TEPID, and
that this is resolved in the \emph{de novo} assembled genome.

To estimate the rate of false negative TE absence calls made using
TEPID, we compared our L\emph{er} TE absence calls to the set of TE
absences in L\emph{er} genome identified previously by aligning
full-length Col-0 TEs to the L\emph{er} reference using BLAT
\cite{Quadrana:2016bi}. We found that 89.6\% (173/193) of these TE
absences were also identified using TEPID, indicating a false negative
rate of around 10\% for TE absence calls. To determine the rate of false
negative TE insertion calls, we ran TEPID using 90 bp paired-end Col-0
reads (Col-0 control samples from \cite{Jiang:2014ih}), aligning reads
to the L\emph{er} PacBio assembly. As TEPID requires a high-quality TE
annotation to discover TE variants, which is not available for the
L\emph{er} assembly, we simply looked for discordant and split read
evidence at the known Col-0-specific TE insertion sites
\cite{Quadrana:2016bi}, and found evidence reaching the TEPID threshold
for a TE insertion call to be made at 89.6\% (173/193) of these sites,
indicating a false negative rate of 10\%. However, we note that this
estimate does not take into account the TEPID refinement step used on
large populations, and so the false negative rate for samples analyzed
in the population from Schmitz et al. (2013) is likely to be slightly
lower than this estimate, as each accession gained on average 4\% more
insertion calls following this refinement step (Figure 2 - figure
supplement \hyperref[fig2s1]{1}).

@@

We used TEPID to analyze previously published 100 bp paired-end genome
resequencing data for 216 different Arabidopsis accessions
\cite{Schmitz:2013iu}, and identified 15,007 TE insertions (Figure 2 -
source data 1) and 8,088 TE absence variants (Figure 2 - source data
2) relative to the Col-0 reference accession, totalling 23,095 unique
TE variants. In most accessions we identified 300-500 TE insertions
(mean = 378) and 1,000-1,500 TE absence variants (mean = 1,279), the
majority of which were shared by two or more accessions (Figure 2 -
figure supplement \hyperref[fig2s2]{2}). PCR validations were
performed for a random subset of 10 insertions and 10 absence variants
in 14 accessions (totalling 280 validations), and confirmed the high
accuracy of TE variant discovery using the TEPID package, with results
similar to that observed using simulated data and the L\emph{er}
genome analysis (Figure 2 - figure supplement
\hyperref[fig2s3]{3}). The number of TE insertions identified was
positively correlated with sequencing depth of coverage, while the
number of TE absence variants identified had no correlation with
sequencing coverage (Figure 2 - figure supplement
\hyperref[fig2s4]{4}A, B), indicating that the sensitivity of TE
absence calls is not limited by sequencing depth, while TE insertion
calls benefitted from high sequencing depth.  However, accessions with
low coverage gained more TE insertion calls during the TEPID
refinement step (Figure 2 - figure supplement \hyperref[fig2s4]{4}C),
indicating that these false negatives were effectively reduced by
leveraging TE variant information for the whole population.

As TE insertion and TE absence calls represent an arbitrary comparison
to the Col-0 reference genome, we sought to remove these arbitrary
comparisons and classify each variant as a new TE insertion or true
deletion of an ancestral TE in the population. To do this, we examined
the minor allele frequency (MAF) of each variant in the population,
under the expectation that the minor allele is the derived allele. We
re-classified common TE absences relative to Col-0 as rare TE
insertions in Col-0, and common TE insertions relative to Col-0 as
true TE deletions in Col-0. Cases where the TE variant was at an high
MAF (\textgreater{}20\%) were assigned NA calls, as it could not be
determined if these were cases where the variant was most likely to be
due to a true TE deletion or due to a new TE insertion. While these
classifications are not definitive, as there will be rare cases where
a true TE deletion has spread through the population and becomes the
common allele, we will correctly classify most TE variants. Overall,
we found 72.3\% of the TE absence variants identified with respect to
the Col-0 reference genome were likely due to a true TE deletion in
these accessions, while 4.8\% were due to new insertions in Col-0 not
shared by others in the population (Table \ref{table2}). Overall, we
identified 15,077 new TE insertions, 5,856 true TE deletions, and
2,162 TE variants at a high MAF that were unable to be classified as
an insertion or deletion (Figure 2 - source data 3).

TE insertions and deletions were distributed throughout chromosome 1 in
a pattern that was similar to the distribution of all Col-0 TEs (Figure
\ref{fig2}A). TE deletions and common TE variants were found in similar
chromosomal regions, as deletion variants represent the rare loss of
common variants. TE deletions and common variants were more highly
enriched in the pericentromeric regions than rare variants or TE
insertions. The distribution of rare TE variants and TE insertions was
somewhat similar to that observed for regions of the genome previously
identified as being differentially methylated in all DNA methylation
contexts (mCG, mCHG, mCHH) between the wild accessions (population
C-DMRs), while population CG-DMRs, differentially methylated in the mCG
context, less frequently overlapped with all types of TE variants
identified \cite{Schmitz:2013iu}. Furthermore, TE variants were depleted
within genes and DNase I hypersensitivity sites \cite{Sullivan:2014ii},
while they were enriched in gene flanking regions and within other
annotated TEs or pseudogenes (Figure \ref{fig2}B). We found that TE deletions and
common TE variants were enriched within the set of TE variants found in
gene bodies (Figure \ref{fig2}C, D). We did not find any significant enrichment
of TE variants within the \emph{KNOT ENGAGED ELEMENT (KEE)} regions,
previously identified as regions that may act as a ``TE sink''
\cite{Grob:2014kg} (Figure 2 - figure supplement \hyperref[fig2s5]{5}). This may indicate
that these regions do not act as a ``TE sink'' as has been previously
proposed, or that the ``TE sink'' activity is restricted to very recent
insertions, as the insertions we analysed in this population were likely
older than those used in the \emph{KEE} study \cite{Grob:2014kg}.

Among the identified TE variants, several TE superfamilies were over-
or under-represented compared to the number expected by chance given
the overall genomic frequency of different TE types (Figure
\ref{fig2}E). In particular, both TE insertions and deletions in the
RC/Helitron superfamily were less numerous than expected, with an
11.5\% depletion of RC/Helitron elements in the set of TE variants. In
contrast, TEs belonging to the LTR/Gypsy superfamily were more
frequently deleted than expected, with a 17\% enrichment in the set of
TE deletions. This was unlikely to be due to a differing ability of
our detection methods to identify TE variants of different lengths, as
the TE variants identified had a similar distribution of lengths as
all Arabidopsis TEs annotated in the Col-0 reference genome (Figure 2
- figure supplement \hyperref[fig2s6]{6}). These enrichments suggest
that the RC/Helitron TEs have been relatively dormant in recent
evolutionary history, while the LTR/Gypsy, which are highly enriched
in the pericentromeric regions, are frequently lost from the
Arabidopsis genome. At the family level, we observed similar patterns
of TE variant enrichment or depletion (Figure 2 - figure supplement
\hyperref[fig2s7]{7}; source data 4).

We further examined Arabidopsis (Col-0) DNA sequencing data from a
transgenerational stress experiment to investigate the possible
minimum number of generations required for TE variants to arise
\cite{Jiang:2014ih}. In one of the three replicates, we identified a
single potential TE insertion in a sample following 10 generations of
single-seed descent under high salinity stress conditions, and no TE
variants in the control single-seed descent set. However, without
experimental validation it remains unclear if this represents a true
variant. Therefore, we conclude that TE variants may arise at a rate
less than 1 insertion in 30 generations under laboratory
conditions. Further experimental work will be required to precisely
determine the rate of transposition in Arabidopsis.

@@

Although thousands of TE variants were identified, they may be linked
(i.e. `tagged') by the previously identified single nucleotide
polymorphisms (SNPs), or unlinked from SNPs across the accessions. We
tested how often common TE variants (\textgreater{}3 \% MAF; 7
accessions) were linked to adjacent SNPs to determine when they would
represent a previously unassessed source of genetic variation between
accessions. SNPs that were previously identified between the
accessions \cite{Schmitz:2013iu} were compared to the presence/absence
of individual TE variants. For the testable TE variants in the
population, the nearest 300 flanking SNPs upstream and downstream of
the TE variant site were analyzed for local linkage disequilibrium
(LD, r\textsuperscript{2}; see methods). TE variants were classified
as being either `low', `mid', or `high' LD variants by comparing
ranked r\textsuperscript{2} values of TE variant to SNPs against the
median ranked r\textsuperscript{2} value for all between SNP
comparisons (SNP-SNP) to account for regional variation in the extent
of SNP-SNP LD (Figure \ref{fig3}A, B) due to recombination rate
variation or selection \cite{Horton:2012jh}. The majority (61\%) of
testable TE variants had low LD with nearby SNPs, and represent a
source of genetic diversity not previously assessed by current
SNP-based genotype calling methods (Figure \ref{fig3}C). 29\% of TE
variants displayed high levels of LD and are tagged by nearby SNPs,
while only 10\% had intermediate levels of LD. We observed a positive
correlation between TE variant MAF and LD state, with variants of a
high minor allele frequency more often classified as high-LD (Figure
\ref{fig3}D). While the proportion of TE variants classified as high,
mid, or low-LD was mostly the same for both TE insertions and TE
deletions, TE variants with a high MAF (\textgreater{}20\%) that were
unable to be classified as either true deletions or as new insertions
had a much higher proportion of high-LD variants (Figure
\ref{fig3}E). This was consistent with the observation that the more
common alleles were more often in a high-LD state. TE variants
displayed a similar distribution over chromosome one regardless of
linkage classification (Figure 3 - figure supplement
\hyperref[fig3s1]{1}). Overall, this analysis revealed an abundance of
previously unexplored genetic variation that exists amongst
Arabidopsis accessions caused by the presence or absence of TEs, and
illustrates the importance of identifying TE variants alongside other
genetic diversity such as SNPs.

@@

To determine whether the TE variants identified affected nearby gene
expression, we compared the steady state transcript abundance within
mature leaf tissue between accessions with and without TE insertions
or deletions, for genes with TE variants located in the 2 kb gene
upstream region, 5' UTR, exon, intron, 3' UTR or 2 kb downstream
region (Figure \ref{fig4}A). While the steady state transcript
abundance of most genes appeared to be unaffected by the presence of a
TE, 196 genes displayed significant differences in transcript
abundance linked with the presence of a TE variant, indicating a role
for these variants in the local regulation of gene expression (1\%
false discovery rate; greater than 2-fold change in transcript
abundance; Figure \ref{fig4}A, Figure 4 - source data 1). We did
not find any functional category enrichments in this set of
differentially expressed genes. As rare TE variants with a MAF less
than 3\% may also be associated with a difference in transcript
abundance, but were unable to be statistically tested due to their
rarity, we performed a burden test for enrichment of rare variants in
the extremes of expression \cite{Zhao:2016gc}. Briefly, this method
counts the frequency of rare variants within each gene expression rank
in the population, and aggregates this information over the entire
population to determine whether an enrichment of rare variants exists
within any gene expression rank. We found a strong enrichment for gene
expression extremes for TE variants in the all gene features tested
(Figure \ref{fig4}B). While TE variants in gene upstream regions
showed a strong enrichment of both high and low gene expression ranks,
TE variants in exons or gene downstream regions seemed to have a
stronger enrichment for low expression ranks than high
ranks. Randomization of the accession names removed these enrichments
completely (Figure 4 - figure supplement \hyperref[fig4s1]{1}), and
there was little difference between TE insertions and TE deletions in
the gene expression rank enrichments found (Figure 4 - figure
supplement \hyperref[fig4s2]{2}). This rare variant analysis further
indicated that TE variants may alter the transcript abundance of
nearby genes.

>>>>

\usepackage{cmbright}
\usepackage{sansmath}
\sansmath
\usepackage[scaled]{helvet}
\renewcommand\familydefault{\sfdefault}
\usepackage[math]{blindtext}

\usepackage[T1]{fontenc}

@@

Variation in the presence or absence of transposable elements (TEs) is a
major source of genetic variation between individuals. Here, we
identified 23,095 TE presence/absence variants between 216 Arabidopsis
accessions. Most TE variants were rare, and we find a burden of rare
variants associated with local extremes of gene expression and DNA
methylation levels within the population. Of the common alleles
identified, two thirds were not in linkage disequilibrium with nearby
SNPs, implicating these variants as a source of novel genetic diversity.
Nearly 200 common TE variants were associated with significantly altered
expression of nearby genes, and a major fraction of inter-accession DNA
methylation differences were associated with nearby TE insertions.
Overall, this demonstrates that TE variants are a rich source of genetic
diversity that likely plays an important role in facilitating epigenomic
and transcriptional differences between individuals, and indicates a
strong genetic basis for epigenetic variation.

@@

Transposable elements (TEs) are mobile genetic elements present in
nearly all studied organisms, and comprise a large fraction of most
eukaryotic genomes. The two classes of TEs are retrotransposons (type I
elements), which transpose via an RNA intermediate requiring a reverse
transcription reaction, and DNA transposons (type II elements), which
transpose via either a cut-paste or, in the case of Helitrons, a rolling
circle mechanism with no RNA intermediate \cite{Wicker:2007en}. TE
activity poses mutagenic potential as a TE insertion may disrupt
functional regions of the genome. Consequently, safeguard mechanisms
have evolved to suppress this activity, including the methylation of
cytosine nucleotides (DNA methylation) to produce 5-methylcytosine (mC),
a modification that can induce transcriptional silencing of the
methylated locus. In \emph{Arabidopsis thaliana }(Arabidopsis), DNA
methylation occurs in all three DNA sequence contexts: mCG, mCHG, and
mCHH, where H is any base but G. Establishment of DNA methylation marks
can be carried out by two distinct pathways -- the RNA-directed DNA
methylation pathway guided by 24 nucleotide (nt) small RNAs (smRNAs),
and the DDM1/CMT2 pathway \cite{Zemach:2013dj, Matzke:2014ek}. A major
function of DNA methylation in Arabidopsis is in the transcriptional
silencing of TEs. Loss of DNA methylation due to mutations in genes
essential for its establishment or maintenance leads to expression of
previously silent TEs, and in some cases transposition
\cite{Mirouze:2009km, Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu, Zemach:2013dj}.

@@

The tools developed previously for identification of novel TE insertion
events have several limitations. They either require a library of active
TE sequences, cannot identify TE absence variants, are not designed with
population studies in mind, or suffer from a high degree of
false-negatives
\cite{Thung:2014ir, Robb:2013gw, Henaff:2015dl, Quadrana:2016bi}. In
order to accurately map the locations of TE presence/absence variants
with respect to a reference genome, we have developed a novel algorithm,
TEPID (Transposable Element Polymorphism IDentification), which is
designed for population studies. We tested our algorithm using both
simulated and real Arabidopsis sequencing data, finding that TEPID is
able to accurately identify TE presence/absence variants with respect to
the Col-0 reference genome. We applied our TE variant identification
method to existing genome resequencing data for 216 different
Arabidopsis accessions \cite{Schmitz:2013iu}, identifying widespread TE
variation amongst these accessions and enabling exploration of TE
diversity and links to gene regulation and epigenomic variation.

@@

We developed TEPID, an analysis pipeline capable of detecting TE
presence/absence variants from paired end DNA sequencing data. TEPID
integrates split and discordant read mapping information, read mapping
quality, sequencing breakpoints, as well as local variations in
sequencing coverage to identify novel TE presence/absence variants with
respect to a reference TE annotation (Figure \ref{fig1}; see methods). This
typically takes 5-10 minutes per accession for Arabidopsis genomic DNA
sequencing data at 20-40x coverage, excluding the read mapping step.
After TE variant discovery has been performed, TEPID then includes a
second refinement step designed for population studies. This examines
each region of the genome where there was a TE insertion identified in
any of the analyzed samples, and checks for evidence of this insertion
in all other samples. In this way, TEPID leverages TE variant
information for a group of related samples to reduce false negative
calls within the group. Testing of TEPID using simulated TE variants in
the Arabidopsis genome showed that it was able to reliably detect
simulated TE variants at sequencing coverage levels commonly used in
genomics studies (Figure 1 - figure supplement \ref{fig1s1}).

In order to further assess the sensitivity and specificity of TE variant
discovery using TEPID, we identified TE variants in the Landsberg
\emph{erecta} (L\emph{er}) accession, and compared these with the
L\emph{er }genome assembly created using long PacBio sequencing reads
\cite{Chin:2013iw}. Previously published 100 bp paired-end L\emph{er}
genome resequencing reads \cite{Schneeberger:2011ft} were first analyzed
using TEPID, enabling identification of 446 TE insertions (Figure 1 -
source data 1) and 758 TE absence variants (Figure 1 - source data 2)
with respect to the Col-0 reference TE annotation. Reads providing
evidence for these variants were then mapped to the L\emph{er} reference
genome, generated by \emph{de novo} assembly using Pacific Biosciences
P5-C3 chemistry with a 20 kb insert library \cite{Chin:2013iw}, using
the same alignment parameters as were used to map reads to the Col-0
reference genome. This resulted in 98.7\% of reads being aligned
concordantly to the L\emph{er} reference, whereas 100\% aligned
discordantly or as split reads to the Col-0 reference genome (Table \ref{table1}).
To find whether reads mapped to homologous regions in both the Col-0 and
L\emph{er} reference genomes, we conducted a blast search
\cite{Camacho:2009fc} using the DNA sequence between read pair mapping
locations in the L\emph{er }genome against the Col-0 genome, and found
the top blast result for 80\% of reads providing evidence for TE
insertions, and 89\% of reads providing evidence for TE absence variants
in L\emph{er}, to be located within 200 bp of the TE variant
reported by TEPID. Thus, reads providing evidence for TE variants map
discordantly or as split reads when mapped to the Col-0 reference
genome, but map concordantly to homologous regions of the L\emph{er} \emph{de
novo} assembled reference genome, indicating that structural variation
is present at the sites identified by TEPID, and that this is resolved
in the \emph{de novo} assembled genome.

To estimate the rate of false negative TE absence calls made using
TEPID, we compared our L\emph{er} TE absence calls to the set of TE
absences in L\emph{er} genome identified previously by aligning
full-length Col-0 TEs to the L\emph{er} reference using BLAT
\cite{Quadrana:2016bi}. We found that 89.6\% (173/193) of these TE
absences were also identified using TEPID, indicating a false negative
rate of \textasciitilde{}10\% for TE absence calls. To determine the
rate of false negative TE insertion calls, we ran TEPID using 90 bp
paired-end Col-0 reads (Col-0 control samples from \cite{Jiang:2014ih}),
aligning reads to the L\emph{er }PacBio assembly. As TEPID requires a
high-quality TE annotation to discover TE variants, which is not
available for the L\emph{er }assembly, we looked for discordant and
split read evidence at the known Col-0-specific TE insertion sites
\cite{Quadrana:2016bi}, and found evidence reaching the TEPID threshold
for a TE insertion call to be made at 89.6\% (173/193) of these sites,
indicating a false negative rate of \textasciitilde{}10\%. However, it should be noted
that this estimate does not take into account the TEPID refinement step
used on large populations, and so the false negative rate for samples
analyzed in the population from Schmitz et al. (2013) is likely to be
lower than this estimate, as each accession gained on average 4\% more
insertion calls following this refinement step (Figure 2 - figure
supplement \hyperref[fig2s1]{1}).

@@

TEPID was used to analyze previously published 100 bp paired-end
genome resequencing data for 216 different Arabidopsis accessions
\cite{Schmitz:2013iu}, and identified 15,007 TE insertions (Figure 2 -
source data 1) and 8,088 TE absence variants (Figure 2 - source data
2) relative to the Col-0 reference accession, totalling 23,095 unique
TE variants. In most accessions TEPID identified 300-500 TE insertions
(mean = 378) and 1,000-1,500 TE absence variants (mean = 1,279), the
majority of which were shared by two or more accessions (Figure 2 -
figure supplement \hyperref[fig2s2]{2}).  PCR validations were
performed for a random subset of 10 insertion and 10 absence variants
in 14 accessions (totalling 280 validations), confirming the high
accuracy of TE variant discovery using the TEPID package, with a false
positive rate for both TE insertion and TE absence identification of
\textasciitilde{}9\%, similar to that observed using simulated data
and the L\emph{er} genome analysis (Figure 2 - figure supplement
\hyperref[fig2s3]{3}). The number of TE insertions identified was
positively correlated with sequencing depth of coverage, while the
number of TE absence variants identified had no correlation with
sequencing coverage (Figure 2 - figure supplement
\hyperref[fig2s4]{4}A, B), indicating that the sensitivity of TE
absence calls is not limited by sequencing depth, while TE insertion
identification benefits from high sequencing depth.  However,
accessions with low coverage gained more TE insertion calls during the
TEPID refinement step (Figure 2 - figure supplement
\hyperref[fig2s4]{4}C), indicating that these false negatives were
effectively reduced by leveraging TE variant information for the whole
population.

As TE insertion and TE absence calls represent an arbitrary comparison
to the Col-0 reference genome, we sought to remove these arbitrary
comparisons and classify each variant as a new TE insertion or true
deletion of an ancestral TE in the population. To do this,the minor
allele frequency (MAF) of each variant in the population was examined,
under the expectation that the minor allele is the derived allele.
Common TE absences relative to Col-0 were re-classified as TE insertions
in Col-0, and common TE insertions relative to Col-0 as true TE
deletions in Col-0. Cases where the TE variant had a high MAF
(\textgreater{}20\%) were assigned NA calls, as it could not be
determined if these were cases where the variant was most likely to be a
true TE deletion or a new TE insertion. While these classifications are
not definitive, as there will be rare cases where a true TE deletion has
spread through the population and becomes the common allele, it will
correctly classify most TE variants. Overall, 72.3\% of the TE absence
variants identified with respect to the Col-0 reference genome were
likely due to a true TE deletion in these accessions, while 4.8\% were
due to insertions in Col-0 not shared by other accessions in the
population (Table \ref{table2}). Overall, we identified 15,077 TE insertions, 5,856
true TE deletions, and 2,162 TE variants at a high MAF that were unable
to be classified as an insertion or deletion (Figure 2 - source data 3).

TE insertions and deletions were distributed throughout chromosome 1
in a pattern that was similar to the distribution of all Col-0 TEs
(Figure \ref{fig2}A). TE deletions and common TE variants were found
in similar chromosomal regions, as deletion variants represent the
rare loss of common variants. TE deletions and common variants were
more highly enriched in the pericentromeric regions than rare variants
or TE insertions. Among TE deletions, type II elements were slightly
less biased towards the centromeres in comparison to the distribution
of type I elements (Figure 2 - figure supplement
\hyperref[fig2s5]{5}). The distribution of rare TE variants and TE
insertions was similar to that observed for regions of the genome
previously identified as being differentially methylated in all DNA
methylation contexts (mCG, mCHG, mCHH) between the wild accessions
(population C-DMRs), while population CG-DMRs, differentially
methylated in the mCG context, less frequently overlapped with all
types of TE variants identified \cite{Schmitz:2013iu}. Furthermore, TE
variants were depleted within genes and DNase I hypersensitivity sites
\cite{Sullivan:2014ii}, while they were enriched in gene flanking
regions and within other annotated TEs or pseudogenes (Figure
\ref{fig2}B). TE deletions and common TE variants were enriched within
the set of TE variants found in gene bodies (Figure 2C, D). No
significant enrichment was found for TE variants within the \emph{KNOT
  ENGAGED ELEMENT (KEE) }regions, previously identified as regions
that may act as a ``TE sink'' \cite{Grob:2014kg} (Figure 2 - figure
supplement \hyperref[fig2s6]{6}). This may indicate that these regions
do not act as a ``TE sink'' as has been previously proposed, or that
the ``TE sink'' activity is restricted to very recent insertions, as
the insertions we analysed in this population were likely older than
those used in the \emph{KEE} study \cite{Grob:2014kg}.

Among the identified TE variants, several TE superfamilies were over-
or under-represented compared to the number expected by chance given
the overall genomic frequency of different TE types (Figure
\ref{fig2}E). In particular, both TE insertions and deletions in the
RC/Helitron superfamily were less numerous than expected, with an
11.5\% depletion of RC/Helitron elements in the set of TE variants. In
contrast, TEs belonging to the LTR/Gypsy superfamily were more
frequently deleted than expected, with a 17\% enrichment in the set of
TE deletions. This was unlikely to be due to a differing ability of
the detection method to identify TE variants of different lengths, as
the TE variants identified had a similar distribution of lengths as
all Arabidopsis TEs annotated in the Col-0 reference genome (Figure 2
- figure supplement \hyperref[fig2s7]{7}). These enrichments suggest
that the RC/Helitron TEs have been relatively dormant in recent
evolutionary history, while the LTR/Gypsy, which are highly enriched
in the pericentromeric regions, are frequently lost from the
Arabidopsis genome. At the family level, we observed similar patterns
of TE variant enrichment or depletion (Figure 2 - figure supplement
\hyperref[fig2s8]{8}; source data 4).

We further examined Arabidopsis (Col-0) DNA sequencing data from a
transgenerational stress experiment to investigate the possible minimum
number of generations required for TE variants to arise
\cite{Jiang:2014ih}. In one of the three replicates subjected to high
salinity stress conditions, we identified a single potential TE
insertion in a sample following 10 generations of single-seed descent,
while no TE variants were identified in any of the three control
single-seed descent replicate sets. However, without experimental
validation it remains unclear if this represents a true variant.
Therefore, we conclude that TE variants may arise at a rate less than 1
insertion in 60 generations under laboratory conditions. Further
experimental work will be required to precisely determine the rate of
transposition in Arabidopsis.

@@

Although thousands of TE variants were identified, they may be linked
to the previously identified single nucleotide polymorphisms (SNPs),
or unlinked from SNPs across the accessions. We tested how frequently
common TE variants (\textgreater{}3 \% MAF, \textgreater{}7
accessions) were linked to adjacent SNPs to determine when they would
represent a previously unassessed source of genetic variation between
accessions.  SNPs that were previously identified between the
accessions \cite{Schmitz:2013iu} were compared to the presence/absence
of individual TE variants.  For the testable TE variants in the
population, the nearest 300 flanking SNPs upstream and 300 SNPs
downstream of the TE variant site were analyzed for local linkage
disequilibrium (LD, r\textsuperscript{2}; see methods). TE variants
were classified as being either `low', `mid', or `high' LD variants by
comparing ranked r\textsuperscript{2} values of TE variant to SNPs
against the median ranked r\textsuperscript{2} value for all between
SNP comparisons (SNP-SNP) to account for regional variation in the
extent of SNP-SNP LD (Figure \ref{fig3}A, B) due to recombination rate
variation or selection \cite{Horton:2012jh}. The majority (61\%) of
testable TE variants had low LD with nearby SNPs, and represent a
source of genetic diversity not previously assessed by SNP-based
genotype calling methods (Figure \ref{fig3}C).  29\% of TE variants
displayed high levels of LD and are tagged by nearby SNPs, while only
10\% had intermediate levels of LD. We observed a positive correlation
between TE variant MAF and LD state, with variants of a high minor
allele frequency more often classified as high-LD (Figure
\ref{fig3}D). While the proportion of TE variants classified as high,
mid, or low-LD was mostly the same for both TE insertions and TE
deletions, TE variants with a high MAF (\textgreater{}20\%) that were
unable to be classified as either true deletions or as new insertions
had a much higher proportion of high-LD variants (Figure \ref{fig3}E).
This was consistent with the observation that the more common alleles
were more often in a high-LD state.  TE variants displayed a similar
distribution over chromosome 1 regardless of linkage classification
(Figure 3 - figure supplement \hyperref[fig3s1]{1}). Overall, this
analysis revealed an abundance of previously uncharacterized genetic
variation that exists amongst Arabidopsis accessions caused by the
presence or absence of TEs, and illustrates the importance of
identifying TE variants alongside other genetic diversity such as
SNPs.

@@

To determine whether the newly discovered TE variants may affect
nearby gene expression, the steady state transcript abundance within
mature leaf tissue was compared between accessions with and without TE
insertions or deletions, for genes with TE variants located in the 2
kb gene upstream region, 5' UTR, exons, introns, 3' UTR or 2 kb
downstream region (Figure \ref{fig4}A). While the steady state
transcript abundance of most genes appeared to be unaffected by the
presence of a TE, 196 genes displayed significant differences in
transcript abundance linked with the presence of a TE variant,
indicating a role for these variants in the local regulation of gene
expression (1\% false discovery rate; \textgreater{}2-fold change in
transcript abundance; Figure \ref{fig4}A, Figure 4 - source data
1). No functional category enrichments in this set of differentially
expressed genes were identified. As rare TE variants with a MAF less
than 3\% may also be associated with a difference in transcript
abundance, but were unable to be statistically tested due to their
rarity, a burden test for enrichment of rare variants in the extremes
of expression was performed \cite{Zhao:2016gc}. Briefly, this method
counts the frequency of rare variants within each gene expression rank
in the population, and aggregates this information over the entire
population to determine whether an enrichment of rare variants exists
within any gene expression rank. A strong enrichment for gene
expression extremes was observed for TE variants in all gene features
tested (Figure \ref{fig4}B). While TE variants in gene upstream
regions showed a strong enrichment of both high and low gene
expression ranks, TE variants in exons or gene downstream regions had
a stronger enrichment for low expression ranks than high
ranks. Randomization of the accession names removed these enrichments
completely (Figure 4 - figure supplement \hyperref[fig4s1]{1}), and
there was little difference between TE insertions and TE deletions in
the gene expression rank enrichments found (Figure 4 - figure
supplement \hyperref[fig4s2]{2}). This rare variant analysis further
indicates that TE variants may alter the transcript abundance of
nearby genes.
